**Urban mobility and socioeconomic deprivation in Latin America after COVID-19**

### 2024-10-23 manuscript/manuscript-231024/manuscript.qmd
transfer

<!--# FR: A few questions to think about: Have we validated the data? i.e. tested their extent of representativeness? Shall we do something quick using census data for each country comparing overall populations?  -->\

@@

![Administrative units within functional urban areas in Argentina, Chile, Colombia and Mexico, coloured according to the average Relative Deprivation Index.](figures/maps.pdf)

@@

![Patterns of recovery for urban mobility in administrative units from Argentina, Chile, Colombia and Mexico.](figures/prediction-rdi-band.pdf)

@@

![Evolution of the parameters characterising the relationship between relative deprivation index and the percentage change in the number of movements relative to the pre-pandemic baseline level of mobility.](figures/regression-evo-nobackground.pdf)

@@

# Methods {#sec-methods}

## Meta-Facebook movement data

@@

@@

# References

>>>>

<!--# FR: A few questions to think about: Have we validated the data? i.e. tested their extent of representativeness? Shall we do something quick using census data for each country comparing overall populations?  -->

\

@@

<!--![Administrative units within functional urban areas in Argentina, Chile, Colombia and Mexico, coloured according to the average Relative Deprivation Index.](figures/maps.pdf)!-->

@@

\<!-
--[Patterns of recovery for urban mobility in administrative units from Argentina, Chile, Colombia and Mexico.](figures/prediction-rdi-band.pdf)
---!\>

@@

<!---![Evolution of the parameters characterising the relationship between relative deprivation index and the percentage change in the number of movements relative to the pre-pandemic baseline level of mobility.](figures/regression-evo-nobackground.pdf)!--->

@@

# Data and Methods {#sec-methods}

## Meta-Facebook data

<!-- Include here population data too !-->

@@

## WorlPpop population data

We used data from WorldPop [@tatem2017] to classify the spatial units of analysis according to their level of urbanisation, and to estimate missing baseline values in the Facebook population data.
WorldPop offers open access gridded population estimates at a resolution as small as 3 arc-seconds approximately 100m and 1km at the Equator, respectively.
WorldPop produces these gridded datasets using top-down (i.e. dissagregating administrative area counts into smaller grid cells) or bottom-up (i.e. interpolating data from counts from sample locations into grid cells) approaches.
For the purposes of this work, we use gridded population data at a resolution of 1km$^2$ in raster format.
We perform a spatial join of the Facebook spatial units (Bing tiles) with the gridded population data and compute the sum of Worldpop populations corresponding to each of the Facebook spatial units.

## Socioeconomic deprivation data

In our analysis, we use the Global Gridded Relative Deprivation Index (GRDI), Version 1 (GRDIv1) data set as a measure of socioeocnomic deprivation.
The GRDI data is made available via NASA's Socioeconomic Data and Applications Centre (SEDAC), at a spatial resolution of 30 arc-seconds, or 1 km$^2$ approximately.
The index quantifies the relative levels of multidimensional deprivation and poverty, where a value of 100 represents the highest level of deprivation and a value of 0 the lowest.
We perform a spatial join of the Facebook spatial units and the gridded relative deprivation data and compute the average RDI corresponding to each of the Facebook spatial units.

## Classification of tiles according to level of urbanisation and socioeconomic deprivation

<!-- Decide whether I'm going for natural breaks or my own splitting. The disadvantage of own splitting is that there is no common logic for all four countries !-->

## Processing Facebook data

To ensure the privacy and anonymity of the users' data, Meta removes information corresponding to data entries where the population or movement count is less than 10 for a specific time or day, retaining only information about the percentage change in the number of counts with respect to the baseline period is retained [@Maas19].
Consequently, we observe many tiles where the population or movement count for either the baseline or crisis period are blank.
<!-- Justify why we do the imputation. Spatial bias (maybe show Moran's I for missing values on a selected day? Or everyday?). Small number bias, underrespresentation of dataentries with low counts that we know for sure exist !-->

### Facebook population data

To imput Facebook population baseline values, we first identify all the baseline values that are available for each spatial unit and weekday (of the three available time windows, we only consider one per day).
We then estimate the missing baseline values for each weekday, based on a linear model for the relationship between the Worldpop population and the Facebook population counts, which are fitted using ordinary least squares.
This is illustrated in Supplementary Figure \ref{fig:imput-baseline-population}.

We then use the complete baseline Facebook population values to compute missing Facebook population counts during the crisis period.
This is possible because, as mentioned above, Meta reports the percentage change in the number of counts with respect to the baseline, even if the counts are not reported due to low value.

### Facebook movement data

The imputation of Facebook movement baseline values is done according to a spatial interaction model (see e.g. [@rowe2022a]).
We considered the population flow between an origin and a destination tile, and model this variable as a function of the Facebook population count at the origin tile on the same weekday, the Facebook population count at the destination on the same weekday and the distance between origin and destination; we also included indicator variables to capture the day of the week and the pair of population density classes corresponding to the origin and destination tiles.
Mathematically, this model can be expressed as \begin{equation}
\mu_{ijw} = \beta_0 + \beta_1pop_{iw} +\beta_2pop_{jw} + \beta_3d_{ij} + \beta_4W + \beta_5C + \varepsilon
\end{equation}

where $\mu_{ijw}$ is the expectation of the flow of people from tile $i$ to tile $j$ on the weekday $w$; $\beta_0$ is an intercept $pop_{iw}$ and $pop_{jw}$ are the Facebook population counts at the origin and destination on weekday $w$ during the baseline period, $d_{ij}$ is the distance between origin and destination, $W$ is a series of indicator variables capturing the day of the week and similarly $C$ is a series of indicator variables reflecting the pair of population density classes for the origin and destination tiles, resulting in X pairs (10 classes × 10 classes minus one so it is not collinear with $\beta_0$), $\beta_{0,1,2,3,4,5}$ are model parameters to be estimated from the observed data.
The error term is denoted by \varepsilon.

To estimate the model parameters, we used a count data regression model.
Specifically, we used a negative binomial regression which is a generalised linear model (GLM) where overdispersion of the error term is assumed, i.e. the variance exceeds the mean.

We compute missing Facebook movement counts during the crisis period by considering the complete Facebook movements baseline and the percentage change in the number of counts with respect to the baseline, which is reported in the Facebook Movements datasets even when the count is not reported due to its low value.

### Correction factors??? For representativeness

## Data analysis

```{=html}
<!-- EXPLAIN - might include the things I had already written below

@@

 !-->
```
# References
